---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

An initial step to load in any packages we need for the data analysis  
- ggplot2 for graphs  

```{r}
library(ggplot2)
```

The first step of the data annlysis is to read in the data from the csv. NOTE: this assumes that the file is located in the current working directory.  
  
There are then a number of processing steps undertaken on the data  
- Firstly, convert the interval values from an integer (5, 15, 100, etc) into time values (0005, 0015, 0100, etc).   
- We do this by adding 4-length of interval variable number of zeros to the start of the value.  
- Then we convert the time into the correct format HH:MM.  
- Finally we paste the date and time together into one variable - datatime.  

```{r}
dataIn <- read.csv("activity.csv")

dataIn$time <- paste0(mapply(function(x,y) paste0(rep(x,y), collapse=""), 0, 4-nchar(dataIn$interval)), dataIn$interval)

dataIn$time <- format(strptime(dataIn$time, format="%H%M"), format="%H:%M")

dataIn$datetime <- paste(dataIn$date, dataIn$time, sep=" ")
```

We'll remove the "NA" values from the data currently. Later in the analysis we will impute values for these entries.  

```{r}
stepError <- is.na(dataIn$steps)
filteredData <- dataIn[!stepError,]
```

## What is mean total number of steps taken per day?

Firstly, lets look at the total number of steps taken in each day. We can use tapply to sum the number steps walked each day. 

```{r}
stepsPerDay <- tapply(filteredData$steps, filteredData$date, sum, simplify = TRUE)
stepsPerDay
```

Looking at the stepsPerDay variable it is clear that there are a number of NA values in the sums - given that the "NA" values have been removed from the intial data it's not clear exactly where these are coming from.

We can the look at the distribution of the total number of steps.   
```{r}
qplot(stepsPerDay, bins=20)
```

The most frequent number of steps seems to be around 10,000 steps a day. Let's look at a quick summary of the data.   

```{r}
summary(stepsPerDay)
```

So, the average number of steps - both the mean and median - is just over 10,700 steps a day. The data shows quite a range of values, between the maximun of 21,000 and the minimum of 41. This lower values seems suspiciously low, and affected by missing data.   

We can also see that there are eight dates that contain NA values.

## What is the average daily activity pattern?

Secondly, let's look at the average pattern of steps over the course of a day.  

```{r}
stepsPerTime <- aggregate(steps ~ time, filteredData, mean)

stepsPerTime$time<- as.factor(stepsPerTime$time)

with(stepsPerTime, plot(time, steps, type="n", xlab="Time", ylab="Number of Steps", main="Steps by Time Interval"))
with(stepsPerTime, lines(time, steps, type="l"))
```

The plot shows the mean number of steps for each 5-minute interval throughout the day. It can be seen that the largest mean number of steps are around 08:30 - 09:00.  There are also peaks around midday and the end of the working day. Unsurprisingly there is very little activity before 06:00 and after 23:00.

The time of day with the maximum mean number of steps is shown below
```{r}
stepsPerTime[stepsPerTime$steps==max(stepsPerTime$steps),]
```

## Imputing missing values

Now we will impute values for the "NA" values that were present in the initial data.   
Fistly, lets look at how many "NA" values there are in the data.

```{r}
totalErrs <- sum(is.na(dataIn$steps))
propErrs <-  round(100*sum(is.na(dataIn$steps))/nrow(dataIn),2)
```

In total there are `r totalErrs` "NA" values in the data. This represents around `r propErrs`% of the entries. 
  
To impute the missing values we will use the mean values for each time interval that we calculated in the previous section of the report. If the steps value is "NA" then we take the steps value for the appropriate time interval in the stepsPerTime dataset.  

```{r}
dataImpute <- dataIn

dataImpute$steps <- ifelse(is.na(dataImpute$steps)==TRUE, stepsPerTime$steps[stepsPerTime$time %in% dataImpute$time], dataImpute$steps)
```

Let's check the top five rows before and after the impution to check it has worked.  
```{r}
head(dataIn)
head(dataImpute)
```

We'll also check the number of "NA" values in the data - which is now zero.

```{r}
sum(is.na(dataImpute$steps))
```

We can now compare the imputed dataset with the initial dataset to see what effect the impution process has had on the results.

```{r}
stepsPerDayImpute <- tapply(dataImpute$steps, dataImpute$date, sum, simplify = TRUE)
qplot(stepsPerDayImpute, bins=20)
```

It can be seen that the distribution of steps has shifted to the right, so the peak is now around 11,000 steps, compared with 10,000 previously.

We can look at the impact of the impution on the mean and median also.   

```{r}
summary(stepsPerDayImpute)
summary(stepsPerDay)
```

Despite the shift in the distribution shown in the histogram, it can be seen that the impution process only has a small effect on the median - increasing it from 10,760 to 10,770. The impution process has no effect on the mean as we're adding in extra data with a total number of steps per day equal to the mean value.   

## Are there differences in activity patterns between weekdays and weekends?

Finally, lets look to see if there is a difference in the number of steps between weekdays and weekends.  

We create a factor variable with two values - "weekday" and "weekend" - by using the weekdays() funtion to get the name of the day of the week.  

```{r}
dataImpute$datetime <- as.Date(dataImpute$datetime)
dataImpute$dayName <- weekdays(dataImpute$datetime)

dataImpute$weekType <- 0
dataImpute$weekType[dataImpute$dayName %in% c("Saturday", "Sunday")] <- 1

dataImpute$weekFactor <- factor(dataImpute$weekType, labels=c("weekday", "weekend"))

head(dataImpute)
```

Now we want to aggregate the number of steps by time interval and the type of day (weekend or weekday) and calculate the mean value for each combination.  

```{r}
stepsPerTimeByFac <- aggregate(steps ~ time+weekFactor, dataImpute, mean)

stepsPerTimeByFac$time<- as.factor(stepsPerTimeByFac$time)
```

We can now plot the data, showing two plots - one for each day type.

```{r}
par(mfrow=c(1,2))

with(stepsPerTimeByFac[stepsPerTimeByFac$weekFactor == "weekday",], plot(time, steps, type="n", xlab="Time", ylab="Number of Steps", main="Weekday: Steps by Time Interval"))
with(stepsPerTimeByFac[stepsPerTimeByFac$weekFactor == "weekday",], lines(time, steps, type="l"))

with(stepsPerTimeByFac[stepsPerTimeByFac$weekFactor == "weekend",], plot(time, steps, type="n", xlab="Time", ylab="Number of Steps", main="Weekend: Steps by Time Interval"))
with(stepsPerTimeByFac[stepsPerTimeByFac$weekFactor == "weekend",], lines(time, steps, type="l"))

```

From these plots it's clear that the number of steps for Weekdays is concentrated early in the morning - presumably a walk to work. There is much less of a clear spkie towards the end of the day - perhaps reflecting a variety of finish times and destinations after work (i.e. shops, gym, pub etc rather than stright home).   

The Weekend data shows a much higher average number of steps throughout the day and no large spike in the data in the morning. The weekend is clearly a more varied pattern with the potential for a number of activities taking place, rather than the repetitive, predictable commute seen in the Weekday data. 
